Manuscript for MADA Project

### 2024-02-22 R/eda-code/eda.qmd
Kevin- Part 2 for Project
EDA plots and updated Manuscript document to include variables of interest.

---
title: "Exploratory Data Analysis: Project Part 2"
author: "Kevin Kosewick & Emma Hardin-Parker"
date: "2/20/2024"
output: html_document
---

@@

```{r}
#Loading necessary libraries
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(tidyverse)
```
# Getting to know the data

@@

```{r}
summary(sleepdata)
structure(sleepdata)
```
The data set contains 373 observations for 13 variables. Some variables are not easily understandable at face value, so we went back to the data source and accessed the data dictionary. The variables are defined as follows:

@@

>>>>

---
title: "Exploratory Data Analysis: Project Part 2"
author: "Kevin Kosewick & Emma Hardin-Parker"
date: "2/20/2024"
output: html_document
editor: 
  markdown: 
    wrap: sentence
---

@@

```{r}
#Loading necessary libraries
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(tidyverse)
```

# Getting to know the data

@@

```{r}
summary(sleepdata)
structure(sleepdata)
```

The data set contains 373 observations for 13 variables.
Some variables are not easily understandable at face value, so we went back to the data source and accessed the data dictionary.
The variables are defined as follows:

@@

# Data Visualization and Exploratory Data Analysis

We will now begin to create some figures to help visualize the relationships in the dataset.
Quality of Sleep is our outcome of interest; we'll explore the effect each other variable seems to have on it.
First, we'll start with the variables that aren't health specific.
I asked Microsoft Copilot in Precise Mode to generate the base code for some plots that I then changed them to be specific to the dataset/have more visual appeal.

```{r}
# Boxplot for Quality of Sleep by Gender
ggplot(sleepdata, aes(x = Gender, y = Quality.of.Sleep)) +
  geom_boxplot() +
  labs(title = "Quality of Sleep by Gender", x = "Gender", y = "Quality of Sleep")

# Create bins for Age in order to create a clean boxplot.
sleepdata$AgeGroup <- cut(sleepdata$Age, breaks = c(20, 30, 40, 50, 60), labels = c("20-30", "30-40", "40-50", "50-60"), include.lowest = TRUE)

# Calculate mean Quality of Sleep for each Age Group to plot with means for cleaner visuals
sleepdata_summary <- sleepdata %>%
  group_by(AgeGroup) %>%
  summarise(MeanQuality = mean(Quality.of.Sleep, na.rm = TRUE))

# Bar plot for Mean Quality of Sleep by Age Group with colorblind-friendly palette
ggplot(sleepdata_summary, aes(x = AgeGroup, y = MeanQuality, fill = AgeGroup)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Mean Quality of Sleep by Age Group", x = "Age Group", y = "Mean Quality of Sleep")

# Calculate mean Quality of Sleep for each Occupation in order to create a bar plot.
sleepdata_summary <- sleepdata %>%
  group_by(Occupation) %>%
  summarise(MeanQuality = mean(Quality.of.Sleep, na.rm = TRUE))

# Bar plot for Mean Quality of Sleep by Occupation with manually specified colors
ggplot(sleepdata_summary, aes(x = Occupation, y = MeanQuality, fill = Occupation)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = rainbow(length(unique(sleepdata_summary$Occupation)))) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Mean Quality of Sleep by Occupation", x = "Occupation", y = "Mean Quality of Sleep")

```

We can see some clear differences in self-reported Quality of Sleep between these different groups already.
Females tend to report an average of 1 higher according to the boxplot.
The different age groups are similar, but the oldest (50-60) reports the highest quality of sleep by far.
Our occupations are a bit all over the place, but Sales Representatives seem to have the worst average scores by far.
We'll now explore Physical Activity and plot it as a predictor for Quality of Sleep.
To make it easier to visualize Physical Activity, we'll create a new variable with categories of Low (0-30 minutes), Medium (30-60), and High (60+).
We'll again use Microsoft Copilot for the basic plot code and modify it to our dataset/preferences.

```{r}
# Histogram for Physical Activity Level
ggplot(sleepdata, aes(x = Physical.Activity.Level)) +
  geom_histogram(binwidth = 10) +
  labs(title = "Histogram of Physical Activity Level", x = "Physical Activity Level (minutes/day)", y = "Count")

# Create a new variable for Physical Activity Level groups. Categorizes as "Low", "Medium", or "High"
sleepdata$PhysicalActivityGroup <- cut(sleepdata$Physical.Activity.Level, breaks = c(0, 30, 60, Inf), labels = c("Low", "Medium", "High"))

# Box plot for Quality of Sleep by Physical Activity Level Group
ggplot(sleepdata, aes(x = PhysicalActivityGroup, y = Quality.of.Sleep)) +
  geom_boxplot() +
  labs(title = "Quality of Sleep by Physical Activity Level Group", x = "Physical Activity Level Group", y = "Quality of Sleep")
```

We can clearly see that Physical Activity level has a positive correlation with Quality of Sleep.
We'll now explore Daily Steps.
We can create a new variable with groups according to the guidelines here at [10,000 steps.org](https://www.10000steps.org.au/articles/healthy-lifestyles/counting-steps/#:~:text=The%20following%20pedometer%20indices%20have%20been%20developed%20to,day%205%20Highly%20active%20is%20more%20than%2012%2C500).
Nobody in the dataset goes over 10000 steps a day, however, so we won't see any data for the "Active" or "Highly Active" groups.

```{r}
# Histogram for Daily Steps
ggplot(sleepdata, aes(x = Daily.Steps)) +
  geom_histogram(binwidth = 500) +
  labs(title = "Histogram of Daily Steps", x = "Daily Steps", y = "Count")

# Create a new variable for Daily Steps groups based off of 10,000 steps.org categories
sleepdata$StepsGroup <- cut(sleepdata$Daily.Steps, 
                       breaks = c(-Inf, 5000, 7500, 10000, 12500, Inf), 
                       labels = c("Sedentary", "Low active", "Somewhat active", "Active", "Highly active"), 
                       include.lowest = TRUE)

# Box plot for Quality of Sleep by Daily Steps Group
ggplot(sleepdata, aes(x = StepsGroup, y = Quality.of.Sleep)) +
  geom_boxplot() +
  labs(title = "Quality of Sleep by Daily Steps Group", x = "Daily Steps Group", y = "Quality of Sleep")

```

Daily Steps don't have as clear of a trend as Physical Activity did, but there do seem to be fewer low Quality of Sleep scores in higher daily steps groups.
Now we'll explore our Sleep Duration variable.
According to [Medical News Today](https://www.medicalnewstoday.com/articles/sleep-cycle-stages#sleep-requirements), adults aged 18-61 are recommended 7+ hours of sleep a night.
We'll keep this in mind when looking at our data.

```{r}
# Histogram for Sleep Duration
ggplot(sleepdata, aes(x = Sleep.Duration)) +
  geom_histogram(binwidth = 0.5) +
  labs(title = "Histogram of Sleep Duration", x = "Sleep Duration (hours)", y = "Count")

# Scatterplot for Quality of Sleep by Sleep Duration
ggplot(sleepdata, aes(x = Sleep.Duration, y = Quality.of.Sleep)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(title = "Quality of Sleep by Sleep Duration", x = "Sleep Duration (hours)", y = "Quality of Sleep")

```

We can see a clear positive correlation between Sleep Duration and Quality of Sleep. This is to be expected. Now we'll begin plotting our final set of variables. We'll start with analyzing the new variable we created with categories of Blood Pressure.

```{r}
# Bar plot for cat_bp
ggplot(sleepdata, aes(x = cat_bp)) +
  geom_bar() +
  labs(title = "Bar Plot of cat_bp", x = "cat_bp", y = "Count")

# Box plot for Quality of Sleep by cat_bp
ggplot(sleepdata, aes(x = cat_bp, y = Quality.of.Sleep)) +
  geom_boxplot() +
  labs(title = "Box Plot of Quality of Sleep by cat_bp", x = "cat_bp", y = "Quality of Sleep")

```
The boxplots look a bit strange for the Elevated and Normal levels, but we can see on the histogram that it's likely due to a shortage of data for these categories. We can see, however, that Quality of Sleep does seem to be negatively correlated with increasing stages of hypertension. Now we'll explore BMI with some plots.

```{r}
# Bar plot for BMI Category
ggplot(sleepdata, aes(x = BMI.Category)) +
  geom_bar() +
  labs(title = "Bar Plot of BMI Category", x = "BMI Category", y = "Count")

# Box plot for Quality of Sleep by BMI Category
ggplot(sleepdata, aes(x = BMI.Category, y = Quality.of.Sleep)) +
  geom_boxplot() +
  labs(title = "Box Plot of Quality of Sleep by BMI Category", x = "BMI Category", y = "Quality of Sleep")

```
We can clearly see that Normal weights seem to have better sleep on average than the obese and overweight categories. However, there are very few in the obese category, so conclusions with that group may not be as supported as others. Now we'll look at Stress Level.

```{r}
# Histogram for Stress Level
ggplot(sleepdata, aes(x = Stress.Level)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Histogram of Stress Level", x = "Stress Level (scale: 1-10)", y = "Count")

# Scatterplot for Quality of Sleep by Stress Level
ggplot(sleepdata, aes(x = Stress.Level, y = Quality.of.Sleep)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(title = "Quality of Sleep by Stress Level", x = "Stress Level (scale: 1-10)", y = "Quality of Sleep")
```
Our scatterplot shows a clear negative correlation between Stress Level and Quality of Sleep. However, this is a subjective scale, so data should be taken with a grain of salt for now. Let's look at Heart Rate related to Quality of Sleep now.

```{r}
# Histogram for Heart Rate
ggplot(sleepdata, aes(x = Heart.Rate)) +
  geom_histogram(binwidth = 5) +
  labs(title = "Histogram of Heart Rate", x = "Heart Rate (bpm)", y = "Count")

# Scatterplot for Quality of Sleep by Heart Rate
ggplot(sleepdata, aes(x = Heart.Rate, y = Quality.of.Sleep)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(title = "Quality of Sleep by Heart Rate", x = "Heart Rate (bpm)", y = "Quality of Sleep")

```

This data is a bit difficult to visualize for now but we can see a negative correlation between the two. Healthy Heart Rates tend to differ between each individual, however, so we'll have to wait until deeper statistical analyses to draw any educated conclusions for now. We've finally reached our last variable: Sleep Disorders.

```{r}
# Bar plot for Sleep Disorder
ggplot(sleepdata, aes(x = Sleep.Disorder)) +
  geom_bar() +
  labs(title = "Bar Plot of Sleep Disorder", x = "Sleep Disorder", y = "Count")

# Box plot for Quality of Sleep by Sleep Disorder
ggplot(sleepdata, aes(x = Sleep.Disorder, y = Quality.of.Sleep)) +
  geom_boxplot() +
  labs(title = "Box Plot of Quality of Sleep by Sleep Disorder", x = "Sleep Disorder", y = "Quality of Sleep")
```
We can see that individuals without a disorder tend to have higher sleep scores than those with Insomnia or Sleep Apnea, which is what we'd expect to see. The dataset also seems to reflect the population accurately, as those without a sleep disorder are more numerous than individuals with one.

### 2024-02-22 products/manuscript/Manuscript.qmd
Kevin- Part 2 for Project
EDA plots and updated Manuscript document to include variables of interest.

>>>>

Project Part 2:

We'll include the data cleaning here in order to generate our plots in this document. Please see the eda.qmd file for full exploratory results and the cleaning process.

```{r}
#load cleaning tools
library(tidyverse)
# Splitting the Blood.Pressure variable into systolic and diastolic variable columns
sleepdata <- sleepdata %>%
  separate(Blood.Pressure, into = c("systolic", "diastolic"), sep = "/")

# Ensuring that the new columns are considered numeric 
sleepdata$systolic <- as.numeric(sleepdata$systolic)
sleepdata$diastolic <- as.numeric(sleepdata$diastolic)

# Creating new categorical variable

sleepdata <- sleepdata %>%
  mutate(cat_bp = case_when(
    systolic < 120 & diastolic < 80 ~ "Normal",
    systolic >= 120 & systolic <= 129 & diastolic < 80 ~ "Elevated",
    (systolic >= 130 & systolic <= 139) | (diastolic >= 80 & diastolic <= 89) ~ "Stage 1 Hypertension",
    systolic >= 140 | diastolic >= 90 ~ "Stage 2 Hypertension"
  ))
# Merging "Normal" and "Normal Weight" attributes into a singular "Normal" category

sleepdata <- sleepdata %>%
  mutate(BMI.Category = case_when(
    BMI.Category %in% c("Normal", "Normal Weight") ~ "Normal",
    BMI.Category == "Overweight" ~ "Overweight",
    BMI.Category == "Obese" ~ "Obese",
  ))
```

Here are the most interesting/distinct plots from the EDA:

Physical Activity plot:

We can clearly see that Physical Activity level has a positive correlation with Quality of Sleep.
```{r}
#load plot package
library(ggplot2)

# Create a new variable for Physical Activity Level groups. Categorizes as "Low", "Medium", or "High"
sleepdata$PhysicalActivityGroup <- cut(sleepdata$Physical.Activity.Level, breaks = c(0, 30, 60, Inf), labels = c("Low", "Medium", "High"))

# Box plot for Quality of Sleep by Physical Activity Level Group
ggplot(sleepdata, aes(x = PhysicalActivityGroup, y = Quality.of.Sleep)) +
  geom_boxplot() +
  labs(title = "Quality of Sleep by Physical Activity Level Group", x = "Physical Activity Level Group", y = "Quality of Sleep")
```

Gender, Age, and Occupation:

We can see some clear differences in self-reported Quality of Sleep between these different groups already.
Females tend to report an average of 1 higher according to the boxplot.
The different age groups are similar, but the oldest (50-60) reports the highest quality of sleep by far.
Our occupations are a bit all over the place, but Sales Representatives seem to have the worst average scores by far.
```{r}
# Boxplot for Quality of Sleep by Gender
ggplot(sleepdata, aes(x = Gender, y = Quality.of.Sleep)) +
  geom_boxplot() +
  labs(title = "Quality of Sleep by Gender", x = "Gender", y = "Quality of Sleep")

# Create bins for Age in order to create a clean boxplot.
sleepdata$AgeGroup <- cut(sleepdata$Age, breaks = c(20, 30, 40, 50, 60), labels = c("20-30", "30-40", "40-50", "50-60"), include.lowest = TRUE)

# Calculate mean Quality of Sleep for each Age Group to plot with means for cleaner visuals
sleepdata_summary <- sleepdata %>%
  group_by(AgeGroup) %>%
  summarise(MeanQuality = mean(Quality.of.Sleep, na.rm = TRUE))

# Bar plot for Mean Quality of Sleep by Age Group with colorblind-friendly palette
ggplot(sleepdata_summary, aes(x = AgeGroup, y = MeanQuality, fill = AgeGroup)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Mean Quality of Sleep by Age Group", x = "Age Group", y = "Mean Quality of Sleep")

# Calculate mean Quality of Sleep for each Occupation in order to create a bar plot.
sleepdata_summary <- sleepdata %>%
  group_by(Occupation) %>%
  summarise(MeanQuality = mean(Quality.of.Sleep, na.rm = TRUE))

# Bar plot for Mean Quality of Sleep by Occupation with manually specified colors
ggplot(sleepdata_summary, aes(x = Occupation, y = MeanQuality, fill = Occupation)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = rainbow(length(unique(sleepdata_summary$Occupation)))) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Mean Quality of Sleep by Occupation", x = "Occupation", y = "Mean Quality of Sleep")
```

BMI:
We can clearly see that Normal weights seem to have better sleep on average than the obese and overweight categories. However, there are very few in the obese category, so conclusions with that group may not be as supported as others.

```{r}
# Box plot for Quality of Sleep by BMI Category
ggplot(sleepdata, aes(x = BMI.Category, y = Quality.of.Sleep)) +
  geom_boxplot() +
  labs(title = "Box Plot of Quality of Sleep by BMI Category", x = "BMI Category", y = "Quality of Sleep")
```

Blood Pressure:

The boxplots look a bit strange for the Elevated and Normal levels, but we can see that it's likely due to a shortage of data for these categories. We can see, however, that Quality of Sleep does seem to be negatively correlated with increasing stages of hypertension.

```{r}
# Bar plot for cat_bp
ggplot(sleepdata, aes(x = cat_bp)) +
  geom_bar() +
  labs(title = "Bar Plot of cat_bp", x = "cat_bp", y = "Count")

# Box plot for Quality of Sleep by cat_bp
ggplot(sleepdata, aes(x = cat_bp, y = Quality.of.Sleep)) +
  geom_boxplot() +
  labs(title = "Box Plot of Quality of Sleep by cat_bp", x = "cat_bp", y = "Quality of Sleep")

```

And Stress Level:
Our scatterplot shows a clear negative correlation between Stress Level and Quality of Sleep. However, this is a subjective scale, so data should be taken with a grain of salt for now.

```{r}
# Scatterplot for Quality of Sleep by Stress Level
ggplot(sleepdata, aes(x = Stress.Level, y = Quality.of.Sleep)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(title = "Quality of Sleep by Stress Level", x = "Stress Level (scale: 1-10)", y = "Quality of Sleep")
```

The other plots were either self explanatory (e.g. Sleep Duration meant higher sleep quality) or had less defined trends (e.g. heart rate and daily steps).

End of Project Part 2
